A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models

### 2024-06-09 manuscript.qmd
work on case study part

To our knowledge, existing approaches for sample size planning for GLMMs have exclusively focused on planning for power. Power analysis methods for multilevel models can be categorized into formula-based methods and simulation-based methods [@murayamaSummarystatisticsbasedPowerAnalysis2022]. Formula-based methods rely on formulas to calculate power directly while simulation-based methods rely on repeatedly simulating data with a known true effect size and estimating power empirically (i.e., what percentage of simulated datasets produces a significant p-value). Available formula-based software packages for multilevel models often do not include GLMMs or are limited to very simple designs [@westfallStatisticalPowerOptimal2014; @murayamaSummarystatisticsbasedPowerAnalysis2022], making it necessary to build data simulations tailored specifically to the study design. A number of tutorials have been published describing how to perform such simulation-based power analysis for multilevel models [@arendStatisticalPowerTwolevel2019; @lafitSelectionNumberParticipants2021; @zimmerSampleSizePlanning2022; @brysbaertPowerAnalysisEffect2018; @kumleEstimatingPowerGeneralized2021; @debruineUnderstandingMixedEffectsModels2021]. However, most of these tutorials focus on linear mixed models (LMMs) and the common study designs [but see @kumleEstimatingPowerGeneralized2021, for a tutorial that also covers more advanced settings]. This narrow focus provides limited guidance for researchers faced with more complex study designs, especially when little prior knowledge about plausible effect sizes is available [see the discussion in @kumleEstimatingPowerGeneralized2021]. Simulation-based power analysis with GLMMs requires making a range of assumptions: The (conditional) distribution assumption specifies the distributional family for the outcome variable. Assumptions about the random effects include the assumption of normality (i.e., that the random effects follow a normal distribution) and the covariance structure among the random effects (i.e., if and how they are correlated). Making these decisions requires understanding the underlying assumptions of the model and ensuring they align with the characteristics of the data being analyzed. Existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, which makes specifying assumptions a practical challenge (see the discussion in [@maxwellSampleSizePlanning2008] and [@kumleEstimatingPowerGeneralized2021]).

- Insert table on existing R packages for power analysis with GLMMs

@@

A theoretical estimand is a quantifiable measure in statistical analysis that is designed to estimate the underlying parameter of interest within a theoretical model or framework. Essentially, it represents a defined characteristic or feature of the theoretical model. This estimand serves as a target for estimation and provides insight into the behavior and properties of the phenomena being studied within the theoretical construct.

@@

When the estimand has been defined, the next step in the research process is to write code that simulates the (assumed) data generating process of the planned study. This requires specifying a generative process for all predictor variables used in the final data analysis. While such assumptions can be quite challenging for observational studies or continuous predictor variables, this is less of a problem for experimental studies with only categorical predictor variables. When all predictor variables have been simulated, one can use the structure of a suitable GLMM to simulate the dependent variable. To simulate the GLMM, one requires plausible values for all model parameters. We will discuss strategies on how these values can be obtained later. Because we have full control over the data generating process in a tailored simulation, it is possible to model specific aspects of the planned study, like data missing by design or assuming that subjects drop out (i.e. deleting their data from the simulated dataset) based on other variables in the simulated dataset. The quality of the results of the sample size planning crucially depends on the plausibility of the simulated data generating process. For this reason, deciding on a specific set of assumptions can easily feel overwhelming. However, we would argue that even a strongly simplified data generating process (e.g. only a small number of interaction effects; only random intercepts and no random slopes; assuming that data is missing completely at random) can yield informative results.

@@

The following R function simulates a full dataset structured according to the design of our case study.
The faux package [@R-faux] contains useful functions when simulating factorial designs, including random effects.

@@

In the first four lines of the function definition, we set some default parameter values (which we will explain in the next section) and load the packages we use to manipulate and simulate data. 
In our case study, each subject (`n_subjects` in total) is assumed to respond to each item (i.e., CT scan; `n_items` in total). Thus, the `add_random` command creates a fully-crossed `data.frame` with `n_subjects` $\times$ `n_items` rows.
We add a between-subject effect with the `add_between` command, simulating that about $25\%$ of subjects are experts.
The next two lines simulate that in $\frac{2}{3}$ of trials, subjects will be presented with AI advice, and if advice is presented, the advice will be correct in about $80\%$ of cases (the variable `advice_correct` is always 0 when no advice is presented). Next, we simulate one random effect for each subject (`u0s`) and for each item (`u0i`).
As assumed by standard GLMMs, the `add_ranef` function draws the random effects from a normal distribution with a mean 0 and a standard deviation specified by the user.
With all design variables done, we are ready to simulate our model equation outlined in the last section.
The linear predictor variable `linpred` ($\eta$ in the GLMM model equations) combines the predictor variables, random effects, and model parameters as assumed by our model.
We then transform the linear predictor with the inverse-link function to compute `y_prob`, the probability that the subject correctly solved the item (in R, the inverse-logit link is computed with `plogis` and the logit link with `qlogis`).
In the final step, we simulate the binary dependent variable `y_bin` (i.e., whether the subject makes an accurate diagnostic decision for the CT scan) by -- for each trial -- drawing from a Bernoulli distribution with success probability `y_prob`.

@@

Tailored simulation-based sample size planning is especially relevant when studies have to be planned in the absence of previous studies with the same design or pilot data. In this scenario, researchers require strategies on how to specify the population parameters used in their data simulation. Population parameters are all model parameters estimated in a GLMM, in particular the regression coefficients of the fixed effects and the standard deviation of the random effects (and the correlation between random effects in more complicated models). In contrast to non-hierarchical linear regression, common heuristics based on standardized effect sizes are less useful or not even available for GLMMs. Our strategies to specify population parameters will require access to domain knowledge from domain experts. Because most domain knowledge of domain experts can only be expressed in unstandardized measurement units of a specific application, we argue that unstandardized effect sizes are usually preferable over standardized effect sizes for tailored simulation-based sample size planning. The basic idea of all strategies is how the data generating process implied by certain values of population parameters can be quantified or visualized in an intuitive way that enables a calibration of population parameters based on the available knowledge of domain experts.

@@

All parameter values in our present case study have been determined based on results from distantly related study designs in the literature. Additionally, we had repeated discussions with our affiliated domain experts in radiology to check whether our assumptions seem plausible.

@@

$$
\begin{aligned}
P(Y=1|\mathbf{X=x}, \mathbf{U} = \mathbf{0}) = \\
= \text{inverse\_logit}[&\beta_0 + \beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s + \\
&\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}]
\end{aligned}
$$
In fact, we would only need the full equation if the subject is an expert and correct advice is presented.
In all other experimental conditions, some terms drop from the equation because they are multiplied by $0$.
The other extreme case would be the probability that a non-expert with average ability solves an item with average difficulty when no advice is presented:
$$
\begin{aligned}
P(Y=1| advice\_present = 0, advice\_correct = 0, expert = 0, u_{0s} = 0, u_{0i} = 0) = \\
= \text{inverse\_logit}[\beta_0]
\end{aligned}
$$
We can revert this perspective by choosing plausible probability values based on domain knowledge and deriving the parameter values implied by these probabilities for each experimental condition.

@@

@tab-probtable shows our set of assumptions concerning the probability that an average subject solves an average item for each experimental condition, as well as the corresponding equations implied by the model.
The table can be used to compute the implied values for the $\beta$ parameters, starting with the first equation and reinserting the computed $\beta$ values in all following equations:

@@

The mathematical structure of GLMMs determines which patterns in data would be produced by the model, if a specific set of values for the population parameters would be specified. The knowledge of how to simulate from a GLMM enables us to compute insightful descriptive statistics that can be compared to available domain knowledge much more easily than the opaque values of model parameters. For example, domain experts might not be able to directly choose plausible values for the  coefficients in a logistic regression model (which are measured on the log-odds scale). However, they should be able to reason about the expected ratio of the binary dependent variable in different experimental conditions, i.e. which descriptive statistics they expect to observe (on average). The job of the analyst who is familiar with the mathematical structure of the GLMM is to produce the model implied value of the insightful descriptive statistic expected by the domain expert. Although insightful descriptive statistics often depend on the model parameters in a non-linear way, it is not necessary to solve the exact relationship mathematically. Instead, one can simply adjust the population parameters by trial and error until the model implied quantities produce the desired result.

@@

For example, we determined the $\beta$ parameters in a way that corresponds to a desired conditional probability of $0.95$, that an expert with average ability solves an item with average difficulty when presented with correct advice (the conditional perspective).
However, even if the model were true, we would not observe that 95% of experts responding to items presented with correct advice from a big sample of subjects drawn from their natural distribution of ability and items drawn from their natural distribution of difficulty (the marginal perspective).
How much the two probabilities differ depends on the standard deviations of the random intercepts (the two probabilities are only equal if both standard deviations would be zero).
We want to use the model implied observed proportion of correct diagnoses in each experimental condition as an insightful descriptive statistics to determine plausible values for the random effect standard deviations.

@@

We tried using these descriptive statistics to judge together with domain experts whether our chosen values for the random effect standard deviations would produce data that align with out domain expertise.
However, although the result was deemed plausible, in our example these statistics are not informative enough to determine a final set of plausible parameter values.
For this reason, we will additionally look at insightful model based quantities.

@@

Because GLMMs are complicated models, descriptive statistics alone are usually not enough to specify plausible values for all model parameters.
This is especially true for the standard deviation of random effects that have complicated (and often unexpected) effects on the model-implied results.
An important advantage of data simulation (where one has full control over parameter values and sample sizes) is that one can produce insightful model based quantities that can never be directly observed in an actual empirical dataset.
For example, in a logistic model with random intercepts for participants, one can produce a visualization of the implied distribution of the probability that a participant on average solves a cognitive task.
Although domain knowledge will probably not suffice to specify this distribution completely, it should be possible to rule out implausible boundary conditions.
For example, the domain expert might deem it implausible that the 5% most able participants have a probability of more than 0.99 to solve the difficult cognitive task.

@@

The discussed inequality of conditional and marginal effects in GLMMs [@fahrmeirRegressionModelsMethods2021] makes their interpretation more difficult.
One must be careful when specifying parameter values based on previous studies or pilot data that use the marginal interpretation (e.g., a pilot study providing an estimate of how often neurologists make an accurate diagnosis based on brain scans). 
However, this does not mean that we cannot use the marginal interpretation (average probability across persons and items) to inform plausible parameter values: When parameter values have been selected, we can compute the implied marginal distributions and compare this information to our domain knowledge.
Then, we can iteratively adjust the parameter values until we are satisfied with the implied distributions.

@@

As an example, @fig-margdist2 reveals a number of implicit assumptions about the comparison between experts and non-experts: With wrong advice, virtually all experts have a higher probability of making a correct diagnosis compared to non-experts when considering only items with average difficulty. In contrast, there is considerable overlap in probability between experts and non-experts with no advice and even higher overlap with correct advice. Patterns like these should be considered carefully and discussed with the domain experts. Parameter values ($\beta$ parameters, and $\sigma_S$) should be adjusted if the implications do not seem reasonable.

We could also have a closer look at variability between items by setting the subject standard deviation to almost zero ($\sigma_S = 0.01$).

@@

The final plot demonstrates that these plots are also useful for spotting standard deviations that are specified too high. For @fig-margdist3 we have set $\sigma_S = 3$ and $\sigma_I = 3$. This implies that in each experimental condition, the probabilities that a subject solves an item are usually close to either 0 or 1, which is not a plausible assumption. However, these high standard deviations do not account for the inherent variability and complexity of human performance. For example, we would expect that a participant with low ability compared to other task experts to solve a difficult item with a probability substantially larger than zero even when presented with wrong advice.

>>>>

To our knowledge, existing approaches for sample size planning for GLMMs have exclusively focused on planning for power. Power analysis methods for multilevel models can be categorized into formula-based methods and simulation-based methods [@murayamaSummarystatisticsbasedPowerAnalysis2022]. Formula-based methods rely on formulas to calculate power directly while simulation-based methods rely on repeatedly simulating data with a known true effect size and estimating power empirically (i.e., what percentage of simulated datasets produces a significant p-value). Available formula-based software packages for multilevel models often do not include GLMMs or are limited to very simple designs [@westfallStatisticalPowerOptimal2014; @murayamaSummarystatisticsbasedPowerAnalysis2022], making it necessary to build data simulations tailored specifically to the study design. A number of tutorials have been published describing how to perform such simulation-based power analysis for multilevel models [@arendStatisticalPowerTwolevel2019; @lafitSelectionNumberParticipants2021; @zimmerSampleSizePlanning2022; @brysbaertPowerAnalysisEffect2018; @kumleEstimatingPowerGeneralized2021; @debruineUnderstandingMixedEffectsModels2021; @kainPracticalGuidePower2015; @greenSIMRPackagePower2016a]. However, many of these tutorials focus on linear mixed models (LMMs) and the common study designs [but see @kumleEstimatingPowerGeneralized2021, for a tutorial that also covers more advanced settings]. This narrow focus provides limited guidance for researchers faced with more complex study designs, especially when little prior knowledge about plausible effect sizes is available [see the discussion in @kumleEstimatingPowerGeneralized2021]. Simulation-based power analysis with GLMMs requires making a range of assumptions: The (conditional) distribution assumption specifies the distributional family for the outcome variable. Assumptions about the random effects include the assumption of normality (i.e., that the random effects follow a normal distribution) and the covariance structure among the random effects (i.e., if and how they are correlated). Making these decisions requires understanding the underlying assumptions of the model and ensuring they align with the characteristics of the data being analyzed. Existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, which makes specifying assumptions a practical challenge (see the discussion in [@maxwellSampleSizePlanning2008] and [@kumleEstimatingPowerGeneralized2021]).

In Table XXX, we give a short review of existing R packages that can be used for power analysis for GLMMs.

@@

@@

When the estimand has been defined, the next step in the research process is to write code that simulates the (assumed) data generating process of the planned study. This requires specifying a generative process for all predictor variables used in the final data analysis. While such assumptions can be quite challenging for observational studies or continuous predictor variables, this is less of a problem for experimental studies with only categorical predictor variables. When all predictor variables have been simulated, one can use the structure of a suitable GLMM to simulate the dependent variable. To simulate the GLMM, one requires plausible values for all model parameters. We will discuss strategies on how these values can be obtained later. Because we have full control over the data generating process in a tailored simulation, it is possible to model specific aspects of the planned study, like data missing by design or assuming that some subjects might drop out. The quality of the results of the sample size planning crucially depends on the plausibility of the simulated data generating process. However, we would argue that even a strongly simplified data generating process (e.g. only a small number of interaction effects; only random intercepts and no random slopes; assuming that data is missing completely at random) can yield informative results.

@@

The following R function simulates a full dataset structured according to the design of our case study.

@@

In the first four lines of the function definition, we set some default parameter values (which we will explain in the next section) and load the packages we use to manipulate and simulate data.^[The faux package [@R-faux] contains useful functions when simulating factorial designs, including random effects.] 
In our case study, each subject (`n_subjects` in total) is assumed to respond to each item (i.e., CT scan; `n_items` in total). Thus, the `add_random` command creates a fully-crossed `data.frame` with `n_subjects` $\times$ `n_items` rows.
We add a between-subject effect with the `add_between` command, simulating that about $25\%$ of subjects are experts.
The next two lines simulate that in $\frac{2}{3}$ of trials, subjects will be presented with AI advice, and if advice is presented, the advice will be correct in about $80\%$ of cases (the variable `advice_correct` is always 0 when no advice is presented). Next, we simulate one random effect for each subject (`u0s`) and for each item (`u0i`).
As assumed by standard GLMMs, the `add_ranef` function draws the random effects from a normal distribution with a mean 0 and a standard deviation specified by the user.
With all design variables done, we are ready to simulate our model equation outlined in the last section.
The linear predictor variable `linpred` ($\eta$ in the GLMM model equations) combines the predictor variables, random effects, and model parameters as assumed by our model.
We then transform the linear predictor with the inverse-link function to compute `y_prob`, the probability that the subject correctly solved the item (in R, the inverse-logit link is computed with `plogis` and the logit link with `qlogis`).
In the final step, we simulate the binary dependent variable `y_bin` (i.e., whether the subject makes an accurate diagnostic decision for the CT scan) by -- for each trial -- drawing from a Bernoulli distribution with success probability `y_prob`.

@@

In the absence of previous studies with the same design or pilot data, researchers require strategies on how to specify the population parameters used in their data simulation. Population parameters are all model parameters estimated in a GLMM, in particular the regression coefficients of the fixed effects and the standard deviation of the random effects (and the correlation between random effects in more complicated models). In contrast to non-hierarchical linear regression, common heuristics based on standardized effect sizes are less useful or not even available for GLMMs. Our strategies to specify population parameters will require access to domain knowledge from domain experts. Because most domain knowledge can only be expressed in unstandardized measurement units of a specific application, we argue that unstandardized effect sizes are usually preferable over standardized effect sizes for tailored simulation-based sample size planning. The basic idea of all strategies is how the data generating process implied by certain values of population parameters can be quantified or visualized in an intuitive way that enables a calibration of population parameters based on the available knowledge of domain experts.

@@

The starting point for all parameter values in our present case study were based on results from distantly related study designs in the literature. Additionally, we had repeated discussions with our affiliated domain experts in radiology to check whether our assumptions seem plausible.

@@

$$
\begin{aligned}
P(Y=1|\mathbf{X=x}, \mathbf{U} = \mathbf{0}) = \\
= \text{inverse\_logit}[&\beta_0 + \beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s + \\
&\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}]
\end{aligned}
$$
In fact, we would only need the full equation if the subject is an expert and correct advice is presented.
In all other experimental conditions, some terms drop from the equation because they are multiplied by $0$.
For example, the probability that a non-expert with average ability solves an item with average difficulty when no advice is presented, only requires the intercept:
$$
\begin{aligned}
P(Y=1| advice\_present = 0, advice\_correct = 0, expert = 0, u_{0s} = 0, u_{0i} = 0) = \\
= \text{inverse\_logit}[\beta_0]
\end{aligned}
$$
We can revert this perspective by choosing plausible probability values based on domain knowledge and deriving the parameter values implied by these probabilities for each experimental condition.

@@

@tab-probtable shows our set of assumptions concerning the probability that an average subject solves an average item for each experimental condition, as well as the corresponding equations implied by the model.
The table can be used to compute the implied values for the $\beta$ parameters, starting with the first equation and reinserting the computed $\beta$ values in all following equations (`b_0` stands for the intercept $\beta_0$, `b_e` for the slope $\beta_e$, and so on):

@@

The mathematical structure of GLMMs determines which patterns in data would be produced by the model, if a specific set of values for the population parameters would be specified. The knowledge of how to simulate from a GLMM enables us to compute insightful descriptive statistics that can be compared to available domain knowledge much more easily than the opaque values of model parameters. For example, domain experts might not be able to directly choose plausible values for the  coefficients in a logistic regression model (which are measured on the log-odds scale). However, they should be able to reason about the expected ratio of the binary dependent variable in different experimental conditions, i.e. which relative frequency they expect to observe (on average). The job of the analyst who is familiar with the mathematical structure of the GLMM is to produce the model implied value of the insightful descriptive statistic expected by the domain expert. Although insightful descriptive statistics usually depend on the model parameters in a non-linear way, it is not necessary to solve the exact relationship mathematically. Instead, one can simply adjust the population parameters by trial and error until the model implied quantities produce the desired result.

@@

For example, we determined the $\beta$ parameters in a way that corresponds to a desired conditional probability of $0.95$, that an expert with average ability solves an item with average difficulty when presented with correct advice (the conditional perspective).
However, even if the model were true, we would not observe that 95% of experts responding to items presented with correct advice from a big sample of subjects drawn from their natural distribution of ability, and items drawn from their natural distribution of difficulty (the marginal perspective).
How much the two probabilities differ depends on the standard deviations of the random intercepts (the two probabilities are only equal if both standard deviations would be zero).
We want to use the model implied observed proportion of correct diagnoses in each experimental condition as an insightful descriptive statistics to determine plausible values for the random effect standard deviations.

@@

We tried using these descriptive statistics to judge together with domain experts whether our chosen values for the random effect standard deviations would produce data that align with out domain expertise.
However, although the result was deemed plausible, these statistics were not informative enough to determine a final set of plausible parameter values.
For this reason, we will additionally look at insightful model based quantities.

@@

Because GLMMs are complicated models, descriptive statistics alone are usually not enough to specify plausible values for all model parameters.
This is especially true for the standard deviation of random effects that have non-linear (and often unexpected) effects on the model-implied results.
An important advantage of data simulation (where one has full control over parameter values and sample sizes) is that one can produce insightful model based quantities that can never be directly observed in an actual empirical dataset.
For example, in a logistic model with random intercepts for participants, one can produce a visualization of the implied distribution of the probability that a participant on average solves a cognitive task.
Although domain knowledge will probably not suffice to specify this distribution completely, it should be possible to rule out implausible boundary conditions.
For example, the domain expert might deem it implausible that the 5% most able participants have a probability of more than 0.99 to solve the difficult cognitive task.

@@

The discussed inequality of conditional and marginal effects in GLMMs [@fahrmeirRegressionModelsMethods2021] makes their interpretation more difficult.
One must be careful when specifying parameter values based on previous studies or pilot data that use the marginal interpretation (e.g., a pilot study providing an estimate of how often radiologists make an accurate diagnosis based on brain scans). 
However, this does not mean that we cannot use the marginal interpretation (average probability across persons and items) to inform plausible parameter values: When parameter values have been selected, we can compute the implied marginal distributions and compare this information to our domain knowledge.
Then, we can iteratively adjust the parameter values until we are satisfied with the implied distributions.

@@

As an example, @fig-margdist2 reveals a number of implicit assumptions about the comparison between experts and non-experts: With wrong advice, virtually all experts have a higher probability of making a correct diagnosis compared to non-experts when considering only items with average difficulty. In contrast, there is considerable overlap in probability between experts and non-experts with no advice and even higher overlap with correct advice. Patterns like these should be considered carefully and discussed with the domain experts. Parameter values ($\beta$ parameters, and $\sigma_S$) should be adjusted if the implications do not seem reasonable.
We could also have a closer look at variability between items by setting the subject standard deviation to almost zero ($\sigma_S = 0.01$).

@@

The final plot demonstrates that these plots are also useful for spotting standard deviations that are specified too high. For @fig-margdist3 we have set $\sigma_S = 3$ and $\sigma_I = 3$. This implies that in each experimental condition, the probabilities that a subject solves an item are usually close to either 0 or 1, which is not a plausible assumption. These high standard deviations do not account for the inherent variability and complexity of human performance. For example, we would expect that a participant with low ability compared to other task experts to solve a difficult item with a probability substantially larger than zero even when presented with wrong advice.
